AAAI Press Formatting Instructions \\for Authors Using \LaTeX{} --- A Guide

### 2021-09-18 publications/iaai21/dora_outlier_detection_iaai.tex
Fill in related work section. #42

\section{Related Work}
In astrophysics, 
    outlier detection methods have been employed for identifying astrophysical
     objects with unique characteristics (Xiong et al., 2010; 
     Storey-Fisher et al., 2020; Hayat et al., 2020) as well as 
     data/modeling artifacts~\citep{wagstaff:des-anom20} (Lochner et al., 2020) in 
     astronomical surveys. Examples of outlier detection applications in 
     Earth science include detecting anomalous objects or materials 
     (Chang and Chiang, 2002; Zhou et al., 2016), data artifacts/noise 
     (Liu et al., 2017; Alvera-Azcárate et al., 2012), change (Touati et al.,
      2020; Zhou et al., 2016),  and ocean extremes (Prochaska et al.). 
      Planetary science applications of outlier detection have mostly focused 
      on prioritizing samples with novel geologic or geochemical features for
       follow-up targeting or analysis (Kerner et al., 2020a; Kerner et al., 
       2020b) \citep{wagstaff:demud13}.

\begin{itemize}
\item Methods for outlier detection
\item Applications of outlier detection related to our focus domains 
(e.g., \citet{kerner2020comparison})
\item Outlier detection toolkits: PyOD Python
package~\citep{zhao:pyod19}, \url{https://github.com/yzhao062/pyod} 
\item Tools for applying outlier detection or other methods across
domains?
\end{itemize}

@@

\subparagraph{Reconstruction error.}
Reconstruction-based methods aim to construct a model of a dataset by learning 
a mapping between the data samples and a lower-dimensional representation that
 minimizes the loss between the input and its reconstruction from the 
 low-dimensional representation~\citep{kerner2020comparison}. The 
 reconstruction error is used to compute the outlier score because samples
  that are unlike the data used to fit the model will be more poorly 
  reconstructed compared to inliers. 
Principal component analysis (PCA) has been used for outlier detection by
scoring samples using the reconstruction error (here, the L2 norm)
between inputs and their inverse
 transformation from the principal subspace~\citep{kerner2020comparison}.
The DEMUD algorithm~\citep{wagstaff:demud13} differs from other
outlier ranking methods: instead of independently scoring all
observations, DEMUD incrementally identifies the most unusual
remaining item, then incorporates it into the model of ``known''
(non-outlier) observations before selecting the next most unusual
item.  DEMUD's goal is to identify diverse outliers and avoid
redundant selections.  Once an outlier is encountered, repeated
occurrences of that outlier are deprioritized.  Methods that score
outliers independently maximize coverage of outliers, while DEMUD
maximizes the fast discovery of distinct outlier types.

\subparagraph{Distance.}
Distance-based methods score outliers based on their distance from some
``background'' which is usually defined by a sample distribution. The 
Reed-Xiaoli (RX) detector is commonly used for anomaly detection in
multispectral and hyperspectral remote sensing; RX uses the Mahalanobis
 distance between a sample and a background dataset mean and covariance
 to score samples~\citep{reed1990adaptive}. The local variant of RX (Local
 RX or LRX) can be used for image or raster data types and scores each pixel in
  an image with respect to a window ``ring'' of pixels surrounding 
  it~\citep{molero2013analysis}. 
  LRX requires two parameters that define the ring: an outer window which 
  defines the size of the window surrounding the pixel and an inner window
   which defines the size of the window around the target pixel to exclude from
   the background distribution.

\subparagraph{Sparsity.}
Sparsity-based methods score outliers based on how isolated or sparse they
are in the feature space. Isolation forests are a common sparsity-based method
that constructs many random binary trees from a 
dataset~\cite{liu2008isolation}. The outlier score for
a sample is quantified as the average distance from the root to the item’s 
leaf. Shorter distances are indicative of outliers because fewer splits were 
required to isolate the sample.

>>>>

\section{Related Work}
Methods for outlier detection have been 
surveyed extensively in prior work
~\cite{pimentel2014review,chandola2009anomaly,markou2003novelty,
markou2003novelty2}.
Methods can be differentiated primarily based on how they define outlierness.
%
Reconstruction-based methods aim to construct a model of a dataset by learning 
a mapping between the data samples and a lower-dimensional representation that
 minimizes the loss between the input and its reconstruction from the 
 low-dimensional representation~\citep{kerner2020comparison}. The 
 reconstruction error is used to compute the outlier score because samples
  that are unlike the data used to fit the model will be more poorly 
  reconstructed compared to inliers. Reconstruction-based methods
  include PCA~\cite{jablonski2015principal}, 
  autoencoders~\cite{richter2017safe,park2018multimodal}, and
  generative adversarial networks~\cite{akcay2018ganomaly}.
%
 Distance-based methods score outliers based on their distance from some
``background'' which can be defined in a variety of ways. For example,
the Reed-Xiaoli (RX) detector computes the 
Mahalanobis distance between each sample and the background dataset
defined by its mean and covariance matrix~\cite{reed1990adaptive}.
%
Sparsity-based methods, such as the isolation forest~\cite{liu2008isolation}
and local outlier factor\cite{breunig2000lof},
 score outliers based on how isolated or sparse they
are in a given feature space. 
%
Probability distribution and density based methods estimate the underlying 
distribution or probability density of a dataset and score samples based on
their likelihood under the learned distribution. Examples include the probabilistic
autoencoder, which scores samples based on the log likelihood under the learned
latent space distribution~\cite{bohm2020probabilistic}, Gaussian mixture 
Models, and kernel density estimators~\cite{chandola2009anomaly}. 
%
Other methods treat detection of outliers as a supervised classification 
problem, usually with only one class constituted by the inliers or 
known normal samples.
Such methods include one-class support vector 
machines~\cite{scholkopf1999support},
support vector data description~\cite{pimentel2014review},
and negative sampling~\cite{sipple:neg-sampling20}.

In astrophysics, 
outlier detection methods have been employed for identifying astrophysical
objects with unique characteristics~\cite{xiong2010anomaly,
storey2020anomaly,hayat2021self} as well as data or modeling 
artifacts in astronomical surveys~\citep{wagstaff:des-anom20,
lochner2021astronomaly}. 
Examples of outlier detection applications in 
Earth science include detecting anomalous objects or materials
~\cite{chang2002anomaly,zhou2016novel}, data artifacts or noise 
~\cite{liu2017unsupervised,alvera2012outlier}, change~\cite{zhou2016novel,
touati2020anomaly}, and ocean extremes~\cite{prochaska2021deep}. 
Planetary science applications of outlier detection have mostly focused 
on prioritizing samples with novel geologic or geochemical features for
follow-up targeting or analysis~\cite{kerner2020analysis,kerner2020comparison,
wagstaff:demud13}.
%
These examples show the promise of applying outlier detection methods in
a variety of real-world science use cases. However, the effort required to 
apply and evaluate the variety of available algorithms is non-trivial and can 
be daunting for non-ML experts, thus impeding the uptake of outlier
detection methods in science applications.  
There is a need for tools that make it easier for domain scientists to apply
outlier detection methods as well as compare results across datasets.
While there have been some efforts to develop tools for facilitating the
application of outlier detection methods~\cite{zhao2019pyod}, they cover
limited data formats and algorithms. The aim of DORA is to fill the need 
for tools that faciliate application, evaluation, and interpretation of outlier
detection methods.

@@

\subparagraph{Reconstruction error.}
Principal component analysis (PCA) has been used for outlier detection by
scoring samples using the reconstruction error (here, the L2 norm)
between inputs and their inverse
 transformation from the principal subspace~\citep{kerner2020comparison}.
The DEMUD algorithm~\citep{wagstaff:demud13} differs from other
outlier ranking methods: instead of independently scoring all
observations, DEMUD incrementally identifies the most unusual
remaining item, then incorporates it into the model of ``known''
(non-outlier) observations before selecting the next most unusual
item.  DEMUD's goal is to identify diverse outliers and avoid
redundant selections.  Once an outlier is encountered, repeated
occurrences of that outlier are deprioritized.  Methods that score
outliers independently maximize coverage of outliers, while DEMUD
maximizes the fast discovery of distinct outlier types.

\subparagraph{Distance.}
 The 
Reed-Xiaoli (RX) detector is commonly used for anomaly detection in
multispectral and hyperspectral remote sensing; RX uses the Mahalanobis
 distance between a sample and a background dataset mean and covariance
 to score samples~\citep{reed1990adaptive}. The local variant of RX (Local
 RX or LRX) can be used for image or raster data types and scores each pixel in
  an image with respect to a window ``ring'' of pixels surrounding 
  it~\citep{molero2013analysis}. 
  LRX requires two parameters that define the ring: an outer window which 
  defines the size of the window surrounding the pixel and an inner window
   which defines the size of the window around the target pixel to exclude from
   the background distribution.

\subparagraph{Sparsity.}
Isolation forests are a common sparsity-based method
that constructs many random binary trees from a 
dataset~\cite{liu2008isolation}. The outlier score for
a sample is quantified as the average distance from the root to the item’s 
leaf. Shorter distances are indicative of outliers because fewer splits were 
required to isolate the sample.
